### 2017-11-29 submission/manuscript.Rmd
refine major questions and problems this manuscript answers in intro

**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study effect and variable region sequenced. Only evenness for adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) resulted in a slightly increased relative risk while lower Shannon diversity and evenness in fecal samples resulted in a significant increase in relative risk for carcinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Using the whole community versus only CRC associated genera to build a prediction model resulted in higher classification success based on Area Under the Curve (AUC) for both adenoma and carcinoma using fecal and tissue samples. For the included studies, most were adequately powered for large effect size differences. This may be more amenable for carcinoma than for adenoma microbiota research due to the smaller community level changes observed in our results.

@@

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma. To accomplish this we first assessed whether the diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma. Next, we assessed how common CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma. Third, using Random Forest models, we analyzed whether the full community or only the CRC associated genera resulted in better model classification based on the area under the curve (AUC). Since the changes in community were subtle for adenoma we also examined what effect and sample size the studies that were used were adequately powered for.

Our analysis found a continuous decrease in Shannon diversity as disease became more severe which correlated with a significantly increased RR for carcinoma. Using only CRC-associated genera, the RR for carcinoma was higher relative to what was observed for the Shannon diversity RR. Conversely, we demonstrate that the AUC of the classification models increased when the full community was incorporated as opposed to only CRC-associated genera. Although we analyzed data sets which sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.

>>>>

**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study effect and variable region sequenced. Only evenness for adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) resulted in a slightly increased relative risk while lower Shannon diversity and evenness in fecal samples resulted in a significant increase in relative risk for carcinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Using the whole community versus only CRC associated genera to build a prediction model resulted in higher classification success based on Area Under the Curve (AUC) for both adenoma and carcinoma using fecal and tissue samples. The most important OTUs for these models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. For the included studies, most were adequately powered for large effect size differences. This may be more amenable for carcinoma than for adenoma microbiota research due to the smaller community level changes observed in our results.

@@

The focus on biomarkers within stool makes sense since the current gold standard for diagnosis is a colonoscopy which can be time-consuming and is not without risks. Stool offers an easy and cost-effective way in which to stratify risk within the population. Although stool represents an easy and less invasive way to assess risk, it is not clear how reflective this sample is to what is actually happening on the adenoma or carcinoma. Some studies have begun to try and assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Since the stool possibly represents an agregate community of the overall gastrointestinal (GI) tract, sampling the adenoma or carinoma directly might do a better job at identifying important microbes and could be drastically different then what is identified in stool. Sampling the tissue directly may provide clearer answers but it is not without problems. Due to bowel prep the communities left for sampling may not be the truly resident microbes but rather a collection of what is able to keep adhered to the mucosa. Further, these samples would contain much more host DNA then microbial DNA. Low biomass samples are very difficult to work with and results can end up being study dependent due to the randomness of contamination [@Salter_contamination_2014]. Due to these variables, one question our meta-analysis aims to answer is whether it is possible to find consistent patterns that emerge across studies regardless of whether they used stool or tissue samples.

This intense focus on identifying biomarkers has identified mostly mouth-associated microbes as potential CRC-associated microbes in the GI tract [@kostic_genomic_2012; @baxter_microbiota-based_2016; @zeller_potential_2014]. The main bacteria of interest arising from this set of microbes has been those within the *Fusobacterium* genus. Yet, the question remains as to whether or not this is indeed the most important genera to be focusing on since many studies microbiota based studies identify a collection of oral microbes rather than single genera [@baxter_microbiota-based_2016; @zeller_potential_2014]. Based on this discrepency, the second question we want to answer with this meta-analysis is if there really is one dominant CRC-associated genera that can be identified across studies.

The heavy focus on the identification of biomarkers has mostly dominated the study of CRC and has had an unattended consequence of reducing the focus on changes that occur within the underlying resident community.  This has been borne out by the majority of previous studies, both within stool and tissue, that have focused predominately on biomarkers as well as the only meta-analysis investigating this area [@shah_leveraging_2017]. Since it has not been investigated in detail the final question that this meta-analysis aims to answer is whether there are consistent detectable community differences as disease severity increases.

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma. To accomplish this we first assessed whether the diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma. Next, we assessed how common CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma. Third, using Random Forest models, we analyzed whether the full community or only the CRC associated genera resulted in better model classification based on the area under the curve (AUC). We observed that the community changes as disease severity worsens and that this community is important for disease classification. However, since the changes in community were subtle for adenoma we also examined what effect and sample size the studies that were used were adequately powered for. Although we analyzed data sets which sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.
